Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition

### 2023-02-16 manuscript/manuscript.qmd
Address comments

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - ref: upf
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - ref: upf
  - name: Ignacio Castillejo
    affiliations:
      - ref: uam
  - name: Núria Sebastian-Galles
    affiliations:
      - ref: upf
affiliations:
  - id: upf
    name: Center for Brain and Cognition, Universitat Pompeu Fabra
  - id: uam
    name: Departamento de Psicología, Universidad Autónoma de Madrid

@@

thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
fig-dpi: 1000
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(ggplot2)
library(patchwork)
library(lubridate)
library(brms)
library(glue)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(scales)
library(bvqdev)

# load and process objects
# load targets as R objects
tar_load(participants)
tar_load(responses)
tar_load(items)
tar_load(bvq_data)
tar_load(model_fit_4)
tar_load(model_log_liks)
tar_load(model_fit_4_prior)
tar_load(model_loos)
tar_load(posterior_draws)
tar_load(marginal_effects_epreds)

items <- bvq_data$pool %>% 
    mutate(item = str_remove(item, "cat_|spa_")) %>% 
    select(item, language, semantic_category, class) %>% 
    right_join(items, by = c("item", "language"))

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure; @bates1994developmental]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Despite the recent increase in the representativeness of the samples in the literature, most studies on early word acquisition have relied exclusively on data from monolingual children while neglecting, to a considerable extent, that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous findings pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals do. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two typologically relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close like Spanish and Catalan (both Romance languages) they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

Floccia et al. pointed to *parallel activation* as a phenomenon that might potentially underpin their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

This account is especially plausible in the light of (1) the evidence pointing to bilingual toddlers' lexicon consisting of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022translation; @pearson1995crosslanguage], and that (2) there is experimental evidence of parallel activation during word comprehension in children and toddlers [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is, to date, scarce nonetheless. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (remarkably larger) sample size. Their results suggest a cognateness effect on vocabulary size, but their study was not aimed at testing the effect of cognateness on age of acquisition at the word-level. Their response variable was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue overcoming some of these pitfalls. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

## A mechanistic account for the cognate facilitation in word acquisition

In this study, we suggest a mechanistic account for the cognate facilitation effect. We build this account from the assumption that bilinguals use fundamentally similar mechanisms than monolinguals to learn words. In both cases, learning a word requires at least two events to take place. First, a representation of the word-form, encoding its segmental information to sufficient detail, must be stored.

This can be accomplished via bottom-up mechanism since early ages. Even if words are embedded in a continuous stream of speech sounds, in which there are no clear boundaries between words, infants can segment discrete word-forms. For instance, they may rely on their familiarity with the typical stress pattern [@jusczyk1995infants], the phonotactic constraints [@friederici1993phonotactic], or the distributional information of their native language [@saffran1996statistical] to detect word boundaries. A second event for word learning to take place is the association between the familiar word form and its referent. This task is far from trivial for many reasons. Referents are rarely labelled ostensively: children are exposed to most instances of a word in referentially ambiguous contexts or in the absence of their referent. Many words are also presented along remarkably variable exemplars of their referent, and even refer to abstract referents such as routines or play interactions (e.g., *uh-oh*; @casey2021moving). Infants nonetheless need few learning instances to create new word-referent associations. The extent to which they do so relying on bottom-up mechanisms such as cross-situational learning [@yu2007rapid], or on constraints and biases that guide their hypotheses about how words and referents are related [@golinkoff1994early], remains a matter of debate. Either way, there is evidence of word-referent links as early as six months of age [@bergelson2012months; @bergelson2015early]. From a connectionist perspective, a word is said to be acquired when the association between its word-form and its conceptualised referent (i.e. semantic representation) has reached some level of associative strength. As infants accumulate experience with word-referent pairings, the association strength between both increases, and this leads to observable events such as infants' showing a stronger preference for a named object compared to unnamed competitors, or their parents being more likely to report such word as acquired in a vocabulary checklist.

Bilinguals deal with a different linguistic landscape than monolinguals, but arguably use similar mechanisms and do so under common constraints [@curtin2011bilingual]. Bilingual word acquisition entails similar events as described above, with the exception that many referents are

they seem to reach their developmental goals of word learning at similar ages than monolinguals. Bilinguals store words

## Two competing mechanisms for a cognate facilitation during word acquisition

We argue that the cognateness facilitation effect takes place

We argue that the facilitative effect of cognateness on word acquisition can be explained by at least two mechanisms, illustrated in @fig-hypotheses. On the one hand, and in line with @mitchell2022cognates, the effect of cognateness might take place during early stages of word acquisition, before stable representations of any of the two word-forms of the translation equivalent has been formed. In this scenario, the auditory recognition of any of the two word-forms might lead to the activation of its (cognate) translation in the other language via their phonological similarity. Since that both word-forms are likely to be heard in the presence of the same referent, the association between one of the word-forms and its referent could be strengthened after the child is exposed to that same word-form, or to its (similar-sounding) phonologically translation.

For instance, in the presence of a tree, an English-Spanish bilingual child might hear *flower* in some occasions, and *flor* in some others. Given the phonological similarity between *flower* and *flor*, it is possible that both word-forms are activated in the child's lexicon every time any of them is encountered. This would lead to an increase in the association strength between each word-form and their common referent every time one of them is encountered, leading to an earlier age of acquisition. The case of a non-cognate translation equivalent, such as *cow* and *vaca* would be different. Although both word-forms are also encountered in the presence of the same referent, their low phonological similarity would not guarantee such facilitation when the child hears either of them, thus leading to a later age of acquisition for both. This account departs from the premise that a word-form can potentially activate its translation equivalent via phonology before acquisition, that is, before the association between such word-form and its referent is consolidated. This would allow both word-forms to mutually boost their acquisition after only few exposures. In this scenario, the effect of cognateness on acquisition should taken place early in age, and should be symmetric for both languages: the facilitation should occur in both word-forms simultaneously.

An alternative account is that cognateness plays a role in the acquisition of a translation pair only when at least one of the word-forms has been acquired, that is, when then association between one of the word-forms and their common referent is consolidated. In this scenario, one of the word-forms can only activate the other via phonological similarity after its representation in the child's lexicon is robust enough. This way, *flower* and *flor* would not benefit from their phonological similarity until wither *flower* or *flor* is acquired. Only after, for instance, *flower* is acquired, the acquisition of *flor* would be boosted by its similarity to *flower*. This account predicts that cognateness has an asymmetric effect on cognate word acquisition: it should be observed only on the acquisition of the word-form of the translation equivalent that was acquired last. As a consequence, and given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014how; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness.

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strenghtened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*) does not undergo the such facilitation. Both word-forms are fairly disimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](img/hypotheses.jpg){#fig-hypotheses}

In line with this second hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  Under the second hypothesis, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

@@

In this study, we investigated the role of cognateness on the acquisition of translation equivalents, and if such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist designed specifically for this study we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We adopted an Bayesian item response theory (*IRT*) approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to the word-form's cognate status, and the participant's frequency of exposure to the word-form (a composite measure of lexical frequency and language degree of exposure), while adjusting for other indices of word difficulty like lexical frequency [@goodman2008does] and length (phonemes), and participant skill like age [see @kachergis2022standard for a similar approach].

>>>>

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
  - name: Ignacio Castillejo
    affiliations:
      - Departamento de Psicología, Universidad Autónoma de Madrid
  - name: Núria Sebastian-Galles
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra

affiliation: Center for Brain and Cognition, Universitat Pompeu Fabra

@@

abstract: |
    Recent studies have reported larger vocabulary sizes for bilingual toddlers learning languages that shared a greater amount of cognates (e.g., English-Dutch). This effect was only observed in the language to which toddlers were exposed the least, and has been suggested to be the result of a cognateness facilitation effect during word acquisition. The specific underpinnings of tyhis effect, and its relationship with the degree of exposure to the language are unclear. In this study, we present vocabulary data from 436 Catalan-Spanish bilinguals and monolinguals aged 12 to 34. We used Bayesian Item Response Theory to estimate participants' probability of acquisition of 604 words, conditional to their word's cognate status and lexical frequency, and the child's age and degree of exposure to each language. We found converging evidence for an earlier age of acquisition for cognate words, mediated by word frequeny and language exposure. Low-frequency words, and words from the language of least exposure benefitted more strongly by their cognate status than high-frequency words. Our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition and processing of words in the other language.

thanks: |
    This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances.

keywords: lexical acquisition, vocabulary, bilingualism, item response theory, bayesian
toc: false
citation: true
fig-dpi: 1000

csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"

format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(ggplot2)
library(patchwork)
library(ggtext)
library(lubridate)
library(brms)
library(glue)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(scales)
library(bvqdev)

# load targets as R objects
tar_load(participants)
tar_load(responses)
tar_load(items)
tar_load(childes)
tar_load(bvq_data)

tar_load(model_fit_4)
tar_load(posterior_draws)

tar_load(model_fit_4_prior)
tar_load(marginal_effects_epreds)
tar_load(posterior_draws_doe)

tar_load(model_log_liks)
tar_load(model_loos)

tar_load(model_rhats)
tar_load(model_neffs)

items <- bvq_data$pool  |>  
    mutate(item = str_remove(item, "cat_|spa_")) |> 
    select(item, language, semantic_category, class) |> 
    right_join(items, by = c("item", "language"))

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [@fenson2007macarthurbates, @samuelson2021precision]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Developmental psychology has increased the representativeness of its samples considerable in recent years, but most literature on early word acquisition relies extensively on data from monolingual children, and neglects to a considerable extent that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous research about bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals do. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology, as it is the case of English and Spanish. It is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results.  The parallel activation hypothesis stems from the language non-selective account of lexical access, and suggests that bilinguals activate both languages simultaneously during speech production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a plausible account for Floccia et al.'s results in the light of (1) the evidence pointing to bilingual toddlers' lexicon consisting of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage], and that (2) there is experimental evidence of parallel activation during word comprehension in children and toddlers [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. It is possible that cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is nonetheless scarce to date. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statiscally more reliable given their remarkably larger sample size. But although their results suggest a cognateness effect on vocabulary size, their study was not aimed at testing the effect of cognateness on age of acquisition at the *word-level.* The response variable in this study was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all words into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

In this study, we provide an account for the cognate facilitation on word acquisition rooted in accumulator models of language acquisition [@hidaka2013computational; @mollica2017how]. We build on the assumption that word acquisition is a function of mainly, but not exclusively, the amount of experience a child has accrued with the word-referent pairing [see @kachergis2022standard for review]. As infants accumulate linguistic input, they also accumulate opportunities to associate familiar word-forms to their corresponding referents. Words with high lexical frequency like *flower* [`r floor(childes[childes$token=="flower",]$freq_million)` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as *headphones* (`r floor(childes[childes$token=="headphones",]$freq_million)` times per million words) [e.g., @goodman2008does]: infants might accumulate learning experiences with *flower* faster than with *headphones*. In the case of bilinguals, the amount of exposures to a word-form an infants receives is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time might need longer to accumulate the same experience with *flower* and *headphones* than an English monolingual does (who listens 100% of the time to English).

We suggest that the cognateness effect proposed by @floccia2018results and @mitchell2022cognates facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *flower* and *flor*, for an English-Spanish bilingual) follow their acquisition trajectory in a similar way they would for a monolingual child: accumulating learning instances, which occur a a rate proportional to the lexical frequency of the word-form and the amount of exposure to the language they belong to. When one of the word-forms is acquired, the acquisition trajectory of the other is impacted, so that the number of needed learning instances left for acquisition are reduced. This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has acquired *flower*, and therefore its form-referent association has been formed, future presentations of the similar sounding translation *flor* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *dog*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014how; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*, which share four phonemes) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strenghtened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*, which do not share any phoneme) does not undergo the such facilitation. Both word-forms are fairly dissimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](img/hypotheses.jpg){#fig-hypotheses}

In line with this hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learnt the English word-form of translation equivalents earlier than the word-form in the additional language. Under this hypothesis, because of children's imbalance in exposure to words in English and in the additional language, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is unlikely to be available yet), while the acquisition of words in the additional language would benefit more frequently from their phonological similarity with the (more likely available) English translation.

@@

In this study, we tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a words *only* once its translation has been already acquired. Using an online vocabulary checklist designed specifically for this study, we collected vocabulary data on production and comprehension from a sample of bilingual children. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian item response theory [*IRT*, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word form, while adjusting for the the number of phonemes in the word form (an indicator of word acquisition difficulty).

We operationalised participants' exposure rate to a word-form using a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it when listening to speech), and  the child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with the participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childes] (see Appendix 1 for more details). Participant's degree of exposure to Catalan and Spanish were reported by families in a language exposure questionnaire they filled before the vocabulary checklist, in which they provided a proportion of the time that the child was estimated to listen to any language they were spoken to regularly. The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.
